Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. At this rate, the af-dist calculation for the target region
was completed in 4.15s (compare to 5.23min for the naive single-threaded
method above).
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: cost of c5.9xlarge per hour * 1.27
% ==                             $4.08 * 1.27
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.

@@

\subsection{Benchmarks}
In this section we describe the methodology used for the simulation-based
benchmarks of Figs~\ref{fig-data-storage},\ref{fig-whole-matrix-compute},
\ref{fig-subset-matrix-compute} and \ref{fig-column-extract}.
The benchmarks use data simulated by conditioning on a large
pedigree of French-Canadians using
\texttt{msprime}~\citep{baumdicker2021efficient},
which have been shown to follow patterns observed in real
data from the same population to a remarkable
degree~\cite{anderson2023on}.
We begin by downloading the simulated ancestral recombination
graph~\cite{brandt2024promise,lewanski2024era,wong2024general}
for chromosome 21 from Zenodo~\cite{anderson2023simulated}
in compressed \texttt{tszip} format. This 552M file
contains the simulated ancestry and mutations for 1.4 million
present-day samples. We then subset the full simulation
down to ${10^1, 10^2, \dots, 10^6}$ samples using
ARG simplification~\cite{kelleher2018efficient,wong2024general},
storing the subsets in \texttt{tskit} format~\cite{tskit2024}.
Note that this procedure captures the growth
in the number of variants (shown in the top x-axis labels)
as we increase sample sizes as a
natural consequence of population-genetic processes.
% jk@holly$ python3 src/collect_data.py site-allele-report
% scaling/data/chr21_10_6.zarr/
% 0 alleles:  100.00%
% 1 alleles:  100.00%
% 2 alleles:  7.92%
% 3 alleles:  0.21%
As a result of simulated mutational processes,
most sites have one alternate allele,
with 7.9\% having two and 0.2\% having three
alternate alleles in the $10^6$ samples dataset.
We then export the variation data from each subset to VCF
using \texttt{tskit vcf subset.ts | bgzip > subset.vcf.gz}
as the starting point for other tools.

@@

\begin{figure}[h]
\includegraphics{figures/s3-throughput}
\caption{Scalability of genotype data processing on AWS.
\label{fig-s3-throughput}}
\end{figure}

\begin{figure}[h]
\includegraphics{figures/s3-network-throughput}
\caption{Chunk data download rate on AWS.
\label{fig-s3-network-throughput}}
\end{figure}

>>>>

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. At this rate, the af-dist calculation for the target 
% 180.80 MB in 4503 files
% Total data: 90060.00 MB in 4503 chnnks
region---172 MiB retrieved in 4503 chunks, 83.87 GiB of genotype data
decompressed and processed---
% NOTE: this doesn't quite up to the 25.5G above:  83.87 / 4.15 = 20.2
was completed in 4.15s.
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: cost of c5.9xlarge per hour * 1.27
% ==                             $4.08 * 1.27
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.

@@

\subsection{Benchmarks}
In this section we describe the methodology used for the simulation-based
benchmarks of Figs~\ref{fig-data-storage},\ref{fig-whole-matrix-compute},
\ref{fig-subset-matrix-compute} and \ref{fig-column-extract}.
The benchmarks use data simulated by conditioning on a large
pedigree of French-Canadians using
\texttt{msprime}~\citep{baumdicker2021efficient},
which have been shown to follow patterns observed in real
data from the same population to a remarkable
degree~\cite{anderson2023on}.
We begin by downloading the simulated ancestral recombination
graph~\cite{brandt2024promise,lewanski2024era,wong2024general}
for chromosome 21 from Zenodo~\cite{anderson2023simulated}
in compressed \texttt{tszip} format. This 552M file
contains the simulated ancestry and mutations for 1.4 million
present-day samples. We then subset the full simulation
down to ${10^1, 10^2, \dots, 10^6}$ samples using
ARG simplification~\cite{kelleher2018efficient,wong2024general},
storing the subsets in \texttt{tskit} format~\cite{ralph2020efficiently,tskit2024}.
Note that this procedure captures the growth
in the number of variants (shown in the top x-axis labels)
as we increase sample sizes as a
natural consequence of population-genetic processes.
% jk@holly$ python3 src/collect_data.py site-allele-report
% scaling/data/chr21_10_6.zarr/
% 0 alleles:  100.00%
% 1 alleles:  100.00%
% 2 alleles:  7.92%
% 3 alleles:  0.21%
As a result of simulated mutational processes,
most sites have one alternate allele,
with 7.9\% having two and 0.2\% having three
alternate alleles in the $10^6$ samples dataset.
We then export the variation data from each subset to VCF
using \texttt{tskit vcf subset.ts | bgzip > subset.vcf.gz}
as the starting point for other tools.

@@

\begin{figure}[h]
\includegraphics{figures/s3-throughput}
\caption{Scalability of genotype data processing on AWS.
Using the prototype parallel Zarr implementation (see Methods) we 
measured the rate genotype data can be decoded to memory (decompression
only) and the rate at which we can perform the full af-dist
calculation as we vary the number of parallel processes.
\label{fig-s3-throughput}}
\end{figure}

\begin{figure}[h]
\includegraphics{figures/s3-network-throughput}
\caption{Chunk data download rate on AWS.
Using the prototype parallel Zarr implementation (see Methods) we 
measured the rate at which chunks of compressed genotype data 
can be downloaded on two different instance types.
\label{fig-s3-network-throughput}}
\end{figure}
